---
title: "wbb"
output: html_document
---

```{r}
library(tidyverse)
```
```{r}
logs <- read.csv("https://dwillis.github.io/sports-data-files/wbbstats1525.csv")

glimpse(logs)
```
```{r}
pergame <- logs |> 
  mutate(
    PointsPerG = ForPoints / Games,
    FGPctPerG = FieldGoalPCT / Games,
    ThreePctPerG = ThreePointPct / Games,
    ReboundsPerG = TotalRebounds/ Games,
    TurnoversPerG = Turnovers / Games) |>
  select(Season, School, Games, PointsPerG, FGPctPerG, ThreePctPerG, ReboundsPerG, TurnoversPerG)
```

```{r}
teamzscores <- pergame |>
  mutate(
    PointsZScore = as.numeric(scale(PointsPerG, center = TRUE, scale = TRUE)),
    FGPctZScore = as.numeric(scale(FGPctPerG, center = TRUE, scale = TRUE)),
    ThreePctZScore= as.numeric(scale(ThreePctPerG, center = TRUE, scale = TRUE)),
    ReboundsZScore = as.numeric(scale(ReboundsPerG, center = TRUE, scale = TRUE)),
    TurnoversZScore = as.numeric(scale(TurnoversPerG, center = TRUE, scale = TRUE)),
    TotalZScore = PointsZScore + FGPctZScore, ThreePctZScore, ReboundsZScore, TurnoversZScore) |>
  arrange(desc(TotalZScore))
```

## Upon running this, we can notice that the data set the season values to the second year in which they were played - i.e., 2020-2021 season was set to 2021. We get a lot of 2021 teams at the top because COVID shortened many teams' seasons - the top initial team, 2021 Duke, played only four games. Weirdly enough, Maryland played 29. I decided to just filter out 2021 seasons because of this. 

```{r}
teamzscores |>
  filter(!Season == "2021")
```
## By doing this, we see that 2017 Maryland was actually the highest performing team over the last decade. Four different UConn teams made it as well, two of which were national championship winning teams, the only two champion teams in the top 10. Some other schools that won the championship within the past decade are on here, but interestingly not the year they won it all - 2017 & 2018 Baylor (won in 2019). Let's take a look at just Maryland.

```{r}
teamzscores |>
  filter(School == "Maryland")
```
## 2021 Maryland is the highest performing Terp team of the last decade by the metric, but also played the least amount of games out of every other year. I'm actually curious as to why they had a full season while others did not. Their 2020 squad was also excellent, but didn't have the chance to compete in the tournament since COVID canceled it. 

### WRITEUP
We often hear the expression "in a league of their own" used to describe great players, coaches, and teams. 

Geno Auriemma and UConn's women's basketball redefines that expression into "a league nobody else might ever reach."

For starters, UConn has won 12 national championships with Auriemma at the helm, including a four-peat from 2013-2016. They also just became national champions again in 2025 - Auriemma's 40th season leading the program. 

Winning a championship has so much that goes into it beyond just performance. The difference with UConn is that they've simply never NOT performed.

Amongst the ten best statistical teams of the past decade - measured by field goal percentage, three point percentage, total rebounds, and total turnovers - Auriemma led four different squads onto that list, two of them national championship teams - the only two on the list. Furthermore, if you look at the greatest teams by field goal percentage each season, UConn has the top four squads, all four in a four year period from 2015-2018.

It's way past time to throw Auriemma into the greatest coach of all time debate. His level of dominance defines a level that nobody else might reach.
