---
title: "Women's Soccer Data"
---

We'll use this notebook to explore some NCAA women's soccer data and [our questions](https://docs.google.com/spreadsheets/d/1G1AjH2iS6kj3kKHCRi0XVy5Lcm-rahJkeTAToVh9J1g/edit?gid=0#gid=0). First, let's load the tidyverse library:

```{r}
library(tidyverse)
```

We'll want to start with 2025 match data, which is located in [this GitHub repository](https://github.com/dwillis/NCAAWomensSoccerData). We can load the match data directly from GitHub using `read_csv`:


```{r}
matches_2025 <- read_csv("https://raw.githubusercontent.com/dwillis/NCAAWomensSoccerData/main/data/ncaa_womens_soccer_matchstats_2025.csv") |> mutate(diff = team_score - opponent_score, total_goals = goals + defensive_goals)
```

Let's take a look at this data to see what it contains, using `head` and `glimpse`:

```{r}
head(matches_2025)
glimpse(matches_2025)
```

Now, onto the questions using aggregates, which means `group_by` and `summarize`:

1. How many shots in total has Maryland attempted and how many have been on goal? Calculate a percentage of shots on goal.

```{r}
matches_2025 |> 
  # filter(team == "Maryland Terrapins, Terps") |> 
  group_by(team) |> 
  summarize(
    total_shots = sum(sh_att)
    shots_on_goal = sum(so_g)
    pct_shots_on_goal = shots_on_goal / total_shots * 100
  )
```

2. What about the same shot stats from last season's first four games? You'll need to load the matches from 2024 - just change the season in the URL - and filter by date to find the first four games.

```{r}
matches_2024 <- read_csv("https://raw.githubusercontent.com/dwillis/NCAAWomensSoccerData/main/data/ncaa_womens_soccer_matchstats_2024.csv") |> mutate(diff = team_score - opponent_score, total_goals = goals + defensive_goals)
```


3. Maryland started 2022 with 4 draws. Have any other teams had as many this season?

```{r}

```


4. How many total corner kicks has Maryland had so far this season and how many have its opponents have? Calculate the difference between the two totals for every team using group_by, summarize and mutate, then filter to find Maryland's totals.

```{r}


```

```{r}
matches_2025 <- read_csv("https://raw.githubusercontent.com/dwillis/NCAAWomensSoccerData/main/data/ncaa_womens_soccer_matchstats_2025.csv") |> mutate(diff = team_score - opponent_score, total_goals = goals + defensive_goals)

matches_2024 <- read_csv("https://raw.githubusercontent.com/dwillis/NCAAWomensSoccerData/main/data/ncaa_womens_soccer_matchstats_2024.csv") |> mutate(diff = team_score - opponent_score, total_goals = goals + defensive_goals)

matches_2024 <- read_csv("https://raw.githubusercontent.com/dwillis/NCAAWomensSoccerData/main/data/ncaa_womens_soccer_matchstats_2024.csv") |> mutate(diff = team_score - opponent_score, total_goals = goals + defensive_goals) |> filter(date < '2024-09-18')
```
t-test

```{r}
t.test(matches_2024$total_goals, matches_2025$total_goals)
```



5. What's the most interesting thing you've seen from these questions?

```{r}
swac_teams <- c("Alabama A&M Bulldogs", "Alabama St. Lady Hornets", "Alcorn Lady Braves", "Ark.-Pine Bluff Lady Lions", "Grambling Lady Tigers", "Jackson St. Lady Tigers", "Mississippi Val. Devilettes", "Prairie View Lady Panthers", "Southern U. Lady Jaguars", "Texas Southern Lady Tigers")


swac_matches <- matches_2025 |> 
  filter(team %in% swac_teams)


big10 <- c("Nebraska Huskers", "Iowa Hawkeyes", "Minnesota Golden Gophers", "Illinois Fighting Illini", "Northwestern Wildcats", "Wisconsin Badgers", "Indiana Hoosiers", "Purdue Boilermakers", "Ohio St. Buckeyes", "Michigan Wolverines", "Michigan St. Spartans", "Penn St. Nittany Lions", "Rutgers Scarlet Knights", "Maryland Terrapins, Terps", "Southern California Trojans", "UCLA Bruins", "Washington Huskies", "Oregon Ducks")

big10_matches <- matches_2025 |> 
  filter(team %in% big10)


matches_2025 |>
  group_by(home_away) |> 
  summarise(
    mean_diff = mean(diff),
    mean_opp_goals = mean(defensive_goals),
    mean_sog = mean(so_g),
    mean_opp_sog = mean(defensive_so_g)
    )

swac_matches |> 
  summarise(
    mean_diff = mean(diff),
    mean_opp_goals = mean(defensive_goals),
    mean_sog = mean(so_g),
    mean_opp_sog = mean(defensive_so_g)
    )

big10_matches |> 
  summarise(
    mean_diff = mean(diff),
    mean_opp_goals = mean(defensive_goals),
    mean_sog = mean(so_g),
    mean_opp_sog = mean(defensive_so_g)
    )

```
## Correlations & Regressions

```{r}
correlations <- read_csv("https://raw.githubusercontent.com/dwillis/NCAAWomensSoccerData/main/data/ncaa_womens_soccer_matchstats_2024.csv")

newcorrelations <- correlations |> 
  mutate(
    differential = goals - defensive_goals
    )

newcorrelations |> summarise(correlation = cor(differential, fouls, method="pearson"))

newcorrelations |> summarise(correlation = cor(differential, yellow_cards, method="pearson"))

fit <- lm(differential ~ fouls, data = newcorrelations)
summary(fit)

newcorrelations |> 
  summarise(correlation = cor(differential, so_g, method="pearson"))

net <- lm(differential ~ so_g, data = newcorrelations)
summary(net)
```

### 2025 Version
```{r}
correlations25 <- read_csv("https://raw.githubusercontent.com/dwillis/NCAAWomensSoccerData/main/data/ncaa_womens_soccer_matchstats_2025.csv")

newcorrelations <- correlations25 |> 
  mutate(
    differential = goals - defensive_goals,
    so_g_diff = so_g - defensive_so_g
    )

newcorrelations |> summarise(correlation = cor(differential, defensive_so_g, method="pearson"))

terps25 <- newcorrelations |> 
  filter(team == "Maryland Terrapins, Terps")

defensive_sog_model <- lm(differential ~ defensive_so_g, data = newcorrelations)
summary(defensive_sog_model)

terps_defensive_sog_model <- lm(differential ~ defensive_so_g, data = terps25)
summary(terps_defensive_sog_model)
```
### Women's Soccer Correlation Story Pitch
The correlation between defensive shots on goal allowed and game outcome for Maryland Women's Soccer is statistically significant. In all four of their losses so far this season, their shots on goal differential has been heavily in the negatives, with each being at least -5. In three of their four losses, the Terps only got one shot on goal. However, what's interesting about this is that in their latest game against Purdue, the Terps were once again in the negatives - -4 - but they won. What specifically did they do to overcome what has typically buried them in past games?
